\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-03-14 honors_thesis/journal_paper/paper.tex
mathy version added with hacked text

%% PSO Stuff
\DeclareMathOperator{\URand}{U}
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{b}_N}}
\providecommand{\pbest}{\ensuremath{\Vec{b}_P}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\coeff}{\ensuremath{\phi}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}

%SpecExPSO Stuff
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  Even in desktop machines, systems with 4 or 8 cores are
commonplace.  In order to efficiently utilize these resources for
computationally intensive problems, algorithms need to run in parallel.

@@

All of these parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new subswarms.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.  However, there are many functions for which performing more
iterations with fewer particles is a better use of resources than performing
fewer iterations with more particles.

@@

@@

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{t+1} &=
		\constriction \left[ \pvel_t +
			\coeff_1\URand()\otimes(\pbest - \ppos_t) +
			\coeff_2\URand()\otimes(\nbest - \ppos_t)
		\right] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_t + \pvel_{t+1}
\end{align}
where \( \URand() \) is a vector of random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen.  \( \coeff_1 \), \(
\coeff_2 \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}.

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel adaptation of the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $n$ neighbors; we use 2 neighbors for all of our
experiments.

@@

In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.  If we can guess
what a particle's next position will be from its current position, we can
evaluate both in parallel and perform two iterations of PSO at once.  Lacking
a good way to guess a particle's next position, we can at least enumerate all
possible next positions of the particle, evaluate all of them at once, and one
of them will be right.

@@

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{cc}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	\hline
	No update&No update\\
	\hline
	No update&Neighbor 1\\
	\hline
	No update&Neighbor 2\\
	\hline
	Updated&No update\\
	\hline
	Updated&Neighbor 1\\
	\hline
	Updated&Neighbor 2\\
	\hline
	Updated&The particle itself\\
	\hline
  \end{tabular}
\end{table}

>>>>

%% PSO Stuff
\DeclareMathOperator*{\argmin}{arg\;min}
\DeclareMathOperator*{\argmax}{arg\;max}
\DeclareMathOperator*{\arginf}{arg\;inf}
\DeclareMathOperator*{\argsup}{arg\;sup}
\providecommand{\pers}{\ensuremath{c}}
\providecommand{\neigh}{\ensuremath{s}}
\providecommand{\nURand}{\ensuremath{U^\neigh}}
\providecommand{\pURand}{\ensuremath{U^\pers}}
\providecommand{\ppos}{\ensuremath{\Vec{x}}}
\providecommand{\pvel}{\ensuremath{\Vec{v}}}
\providecommand{\nbest}{\ensuremath{\Vec{x}^{\neigh*}}}
\providecommand{\pbest}{\ensuremath{\Vec{x}^{\pers*}}}
\providecommand{\constriction}{\ensuremath{\chi}}
\providecommand{\ncoeff}{\ensuremath{\phi^s}}
\providecommand{\pcoeff}{\ensuremath{\phi^P}}
\providecommand{\obs}{\ensuremath{\Vec{\xi}}}
\providecommand{\ofunc}{\ensuremath{f}}
\providecommand{\swarm}{\ensuremath{swarm}}

%SpecExPSO Stuff
\providecommand{\indic}{\ensuremath{I}}
\providecommand{\specvel}{\ensuremath{V}}
\providecommand{\specpos}{\ensuremath{X}}
\providecommand{\ln}{\ensuremath{ln}}
\providecommand{\rn}{\ensuremath{rn}}
\providecommand{\caseset}{\ensuremath{\mathcal{C}}}
\providecommand{\casedef}{\ensuremath{(\pbest:\nbest)}}
\providecommand{\casexn}{\ensuremath{(x:\neigh*)}}
\providecommand{\casexx}{\ensuremath{(x:x)}}
\providecommand{\casexl}{\ensuremath{(x:\ln)}}
\providecommand{\casexr}{\ensuremath{(x:\rn)}}
\providecommand{\casepn}{\ensuremath{(\pers*:\neigh*)}}
\providecommand{\casepl}{\ensuremath{(\pers*:\ln)}}
\providecommand{\casepr}{\ensuremath{(\pers*:\rn)}}
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

Particle swarm optimization (PSO) has been found to be a highly robust and
effective algorithm for solving many types of optimization problems.  For much
of the algorithm's history, PSO was run serially on a single machine.  However,
the world's computing power is increasingly coming from large clusters of
processors.  i
%Even in desktop machines, systems with 4 or 8 cores are
%commonplace.  
In order to efficiently utilize these resources for
computationally intensive problems, algorithms need to run in parallel.

@@

The purpose of this paper is to consider pso parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take on the order of hours.
Our purpose is to explore the question ``since pso is often run with
a swarm of size of 100, what would one do if 500 or a few thousand processors were available?''

Previous parallel techniques use additional processors to increase the
number of particles in the swarm, either by adding individual particles or by
adding entire new sub-swarms but in all cases increasing the number of particles.  The number of iterations of PSO that the
algorithms can perform is thus inherently limited by the time it takes to
evaluate the objective function---additional processors add more particles, not
more iterations.
for many functions there comes a point of diminishing returns with respect to adding particles.
>>>>>>>>> TODO WARNING WARNING MADE UP NUMBERS GET REAL ONES!! <<<<<<<<<<<<<<
For the well-know benchmark function Sphere and as shown in Figure XXXXa increasing the swarm size from 20 to 120 has a significant effect on the 
speed of convergence. In this case, the swam goes below $1*10^5$ in x iteration with 20 particles, but 
achieves this same value in  y iterations with a swarms size of 120, the additional 100 particles.
However consider the change going from 2000 particles to 2100 particles. Here the additional 100 particles
reduce the number of iterations to achieve $1*10^5$ from x iterations to y.
Other benchmark function behave similarly. ans shown in fig xxxxb for ?? and xxxxc ?? (in all case 20 dimensions, average of 50 runs).
Certainly these benchmark functions do not prove that this will be the case for all functions,
but it does provide some evidence that for at least some functions, adding additional particles does not seem to have much benefit
after some point, some time earlier fig xx some time later. This phenomena has been commented on by others
in the form of recommended swarms sizes for various types of functions cite standard swarm paper.
\begin{quote}
Skip this??

However, there are many functions for which performing more
iterations with fewer particles is a better use of resources than performing
fewer iterations with more particles.  
\end{quote}

@@

\begin{quote}
Skip this??

@@

But the following does need some intro
\end{quote}

@@

At each iteration of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle $i$ are updated as follows:
\begin{align}
\label{eq:velupdate}
	\pvel_{i,t+1} &=
		\constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
		\right] \\
\label{eq:posupdate}
	\ppos_{i,t+1} &= \ppos_{i,t} + \pvel_{i,t+1}
\end{align}
where \( \nURand_{i,t} \) and \( \nURand_{i,t} \) are vectors of i.i.d. random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we have made the particle identifier $i$
and the time step $t$ or $t+1$ explicit in the subscript and moved other qualifiers,
such as the personal $P$ and personal best $P*$ qualifiers to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel surrogate for the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $n$ neighbors; we use 2 neighbors for all of our
experiments.
%TODO except as noted??

@@

In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.  If we can guess
what a particle's next position will be from its current position, we can
evaluate both in parallel and perform two iterations of PSO at once.  Lacking
a good way to guess a particle's next position, we can at least enumerate all
possible next positions of the particle, evaluate all of them at once, and one
of them would have been computed by normal PSO.

We do this as described here.

First consider formula for the \emph{next} iteration of PSO, which is identical to 
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} except the time subscripts are incremented.
As noted above and for simplicity in this explanation, we restrict our discussion here to
the case of a Ring topology with two neighbors which we will call the
``right neighbor'' and ``left neighbor'' neighbors identified for each
particle $i$ as $\rn(i)$ and $\ln(i)$ respectively.

Next note that the computation of $\pbest$ and $\nbest$ is not seen in
the update equations~\eqref{eq:velupdate} and \eqref{eq:posupdate}, so
we will make this mathematically explicit.
This update is usually implemented as a pair of independent ``if``-statements, but that is
a bit awkward for our purpose here so we instead use a set of indicator variables,
one each representing each possible update to the \emph{pair} $\casedef$.
These indicator variables allow us to easily identify the cases in which updates will
be needed for $\nbest$, $\pbest$ or both.

For example, we can write $\casexn$ to identify the case where $\pbest$ will
be updated with the better position $\ppos_{i,t}$, shown in the first positions,
but a no new neighborhood best was found so
$\nbest$ remains in the second position.

For any topology with 2 neighbors there are 7 possible update cases listed in
Table~\ref{tab:evals}.
This list assumes, as is typically the case, that each particle is also
part of its own neighborhood (can cause an update to $\nbest$). The list also
omits the case where $\nbest$ is updated by $\ppos_{i,t}$ but $\pbest$ was not updated, as this is
logically impossible. $\caseset$ represents the set of all cases.

For minimization, we can define the indicator variable representing the case where for particle $i$ at time $t$
$\pbest$ will
be updated with the better position $x_{i,t}$ but
$\nbest$ will not updated by:
\begin{align}
\nonumber
	\indic_{i,t}^{\casexn} & (\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))= \\
\nonumber
	& 1 \ if \  \ofunc(\ppos_{i,t+1}) > \ofunc(\pbest_{i,t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{i,t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\ln(i),t}) \\
\nonumber
	& \quad \quad \ \wedge \ \ofunc(\nbest_{i,t}) > \ofunc(\ppos_{\rn(i),t}) \\
\label{eq:deficasexn}
	& 0 \ otherwise
\end{align}

For compactness in our equations we represent this without the parameters since
they can be inferred from the subscripts on $\indic_{i,t}^{\casexn}$.
The indicators variables for the other update cases shown in 
Table~\ref{tab:evals}
($\casepn$, $\casepl$, $\casepr$, etc.)
can be defined in similar fashion but obviously using the
corresponding boolean expression.

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a velocity update equation; once again using $\casexn$ as an example:
\begin{align}
\nonumber
	\specvel_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
\label{eq:defvocasexn}
		& \constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t}\otimes(\pbest_{i,t} - \ppos_{i,t}) +
			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
		\right]
\end{align}

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{i,t}^{\casexn}$.
Since this equation is restricted to a specific case, $\casexn$ for this example,
it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
$\pbest_{i,t+1}=\ppos_{i,t+1}$ and $\nbest_{i,t+1}=\nbest_{i,t+1}$ we can re-write
$\specvel_{i,t}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{i,t+1}^{\casexn} & = \\
\label{eq:defvcasexn}
		& \constriction \left[ \pvel_t +
			\pcoeff\pURand_{i,t}\otimes(\ppos_{i,t} - \ppos_{i,t}) +
			\ncoeff\nURand_{i,t}\otimes(\nbest_{i,t} - \ppos_{i,t})
		\right]
\end{align}

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a position update equation with its compact form; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{i,t+1}^{\casexn} & (\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})= \\
\nonumber
	\specpos_{i,t+1}^{\casexn} & =  \ppos_{i,t} + \specvel_{i,t+1}^{\casexn}
\end{align}

With this notation we can re-write the ``next iteration'' form of the PSO motion equations as:
\begin{align}
\label{eq:vel2update}
	\pvel_{i,t+2} &= \sum_{c \in \caseset} \ \specvel_{i,t+1}^{c} \ \indic_{i,t}^{c} \\
\label{eq:pos2update}
	\ppos_{i,t+2} &= \sum_{c \in \caseset} \ \specpos_{i,t+1}^{c} \ \indic_{i,t}^{c}
\end{align}

The value of the objective function at $\ppos_{i,t+2}$ is given by:
\begin{align}
\label{eq:pos2value}
	\ofunc (\ppos_{i,t+2}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{i,t+1}^{c}) \ \indic_{i,t}^{c}
\end{align}
or, with $\specpos$ and $\indic$ written with the parameters which were omitted above:
\begin{align}
\nonumber
\ofunc & (\ppos_{i,t+2}) = \sum_{c \in \caseset} \\
\nonumber
& \ofunc(\specpos_{i,t+1}^{c}(\ppos_{i,t+1} ,\ppos_{\ln(i),t+1},\ppos_{\rn(i),t+1} ,\pbest_{i,t} ,\nbest_{i,t})) \\
\label{eq:pos2valuelong}
& \indic_{i,t}^{c}(\ofunc ( \ppos_{i,t+1} ) ,\ofunc(\ppos_{\ln(i),t+1}),\ofunc(\ppos_{\rn(i),t+1}) ,\ofunc(\pbest_{i,t}) ,\ofunc(\nbest_{i,t}))
\end{align}

None of the $\specpos_{i,t+1}^{c} \ \ c \in\ \caseset$ depend upon $f(i,x_t)$
or any other objective function evaluation at $t+1$.
Thus PSO has been refactored such that the algorithm can begin computing
all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.

This refactoring comes a significant expense, instead of just computing
$\ofunc(\ppos_{i,t+1})$ and then $\ofunc(\ppos_{i,t+2})$ in the usual way, this refactored PSO
can compute $\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time.
Once all of the evaluations $f(\ppos_{i,t+1})$ are completed only one
of the indicator functions $\indic_{i,t}^{c} \ \ c \in \caseset$ will have been set to one;
the rest will be ignored, and might just as well never have been computed.
We call these evaluated points $\specpos_{i,t+1}^{c}$ ``speculative children''
because they may or may not be needed.

The potential value of these speculative children lies in the phase used above
``this refactored PSO
can compute $\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
Remember that our intent is to accelerate parallel PSO and that we
are interested in objective functions that have at least non-trivial computational
complexity.
Thus we assume that computing
$\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ is the computationally difficult part.

Suppose that unlimited processors were available and that the evaluation of 
the objective function took 1 hour. If for each particle $i$ 
the computations
$\ofunc(\ppos_{i,t+1})$ and all 
of the $\ofunc(\specpos_{i,t+1}^{c}) \ \ c \in \caseset$ were sent to separate
processors and completed in parallel.
In this environment all of the function evaluations needed for 2 iterations of PSO
could be completed in just one hour where as the conventional implementation would
require two hours for those same 2 iterations of PSO.

the same time''. 
in the indicator functions $indic_{i,t}^{c} \ \ c \in \caseset$
given the values $\ofunc(\ppos_{i,t+1}) i \in \swarm$ 
is inconsequential relative to the computational cost of obtaining those values 
$\ofunc(\ppos_{i,t+1})$ and all

@@

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{lcc}
	Identifier&$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	\hline
	$\casepn$&No update&No update\\
	\hline
	$\casepl$&No update&Left Neighbor\\
	\hline
	$\casepr$&No update&Right Neighbor\\
	\hline
	$\casexn$&Updated&No update\\
	\hline
	$\casexl$&Updated&Left Neighbor\\
	\hline
	$\casexr$&Updated&Right Neighbor\\
	\hline
	$\casexx$&Updated&The particle itself\\
	\hline
  \end{tabular}
\end{table}
